\PYGEOMETRY{}: a PYTHON library for the creation of Monte Carlo radiation transport physical geometries

### 2020-09-04 public/paper/pyg4ometry.tex
Paper update
Paper update (following meeting with SW 4/10/20)

There are existing codes that have functionality similar to \PYGEOMETRY{}. Tools exist to convert CAD files to GDML, open source examples are GUIMesh \cite{GUIMesh_Pinto}, commercial products include ESABASE2 \cite{ESABASE2} and FASTRAD \cite{FASTRAD}. There are also tools from the fusion and 
neutronics community which  can convert CAD geometry into formats usable by RTMC codes, examples include DAGMC \cite{DAGMC} and McCad \cite{McCad}.  
In principle CAD software can export shape data to STL (or other similar mesh formats), which can be used by Geant4 \cite{poole2012acad}. In practice using a lot of CAD models is difficult if a CAD model comprises of a large number of parts, exporting, assigning material to and placing the STL components into the RTMC can be a 
cumbersome task. The existent set of software does not provide a complete set of tools to efficiently create complex geometries.

@@

%%%%%%%%%%%%%%%%% Geant4 geometery
Geant4 geometry description is a mixture of BREP, CSG and tessellated concepts. Geant4 includes XX fundamental solids, although it does not store a sense 
of topology present in traditional CAD BREP systems. One of the fundamental solids is a {\em TessellatedSolid} which can used to represent STL or PLY files. 
Geant4 also provides the ability to perform Boolean operations on these primitive solids. The richest and most flexible geometry description is currently used by
Geant4. Not only do solid objects need to be defined but also placed in a world coordinate system. Geant4 has two concepts which facilitate this {\it Logical Volumes} and {\it Physical Volumes}. A {\it Logical Volume} is a region of space  that is defined by an outer solid but also other attributes like material, magnetic field 
and zero or more physical {\it daughter} volumes. A Physical Volume is a placement (or instance) of a Logical Volume.

@@

\subsection{Tessellation of solids (meshing)}
For each Geant4 primitive solid a triangular tessellated Vertex-Face mesh is generated and cached. This mesh is then used to determine the extent of placed 
instances of geometry (physical volumes) and meshes for CSG derived solids. CSG mesh calculations are performed using a BSP tree technique in pure PYTHON 
\cite{pycsg} or the Computational Geometry Algorithms Library (CGAL) surface meshes \cite{cgal:bsmf-sm-20b}. In general the CGAL implementation is one or two orders of magnitude faster than the pure-PYTHON CSG implementation and must be used for large geometries. Triangular meshes based on CSG 
operations involving curved surfaces often contain large numbers of thin triangles. Before meshes are visualised or written to disk various polygon mesh 
algorithms  from CGAL \cite{cgal:lty-pmp-20b} can be employed to give the meshes more desirable features.

@@

\subsection{Visualisation} \label{sec:visualisation}
A  geometry hierarchy can be viewed using the popular Visualisation Toolkit (VTK). No separate scene graph is required as the Geant4 volume hierarchy is sufficient 
to place the meshes associated with each physical volume. A daughter volume is placed within a logical volume with a rotation $\mathbf{R}_d$, scale $\mathbf{S}_d$ and translation $\mathbf{T}_d$.

@@

\subsection{Overlap detection}
All RTMC codes can not handle spatial overlap between two geometric objects and likely are to have ill-defined behaviour when tracking particles  
through such a situation.  A key feature of \PYGEOMETRY{} is the detection of potential overlaps efficiently, it does this by performing an intersection operation between solid instances and determining if the resulting mesh is empty. Figure~\ref{fig:overlap} shows the potential overlaps, a) protrusion of a daughter from the mother, b)  finite volume intersection between two daughters  and c) an overlap where two daughters share a face. 
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=8cm]{./diagrams/overlap.pdf}
\caption{Schematic of the three different types of overlap between daughters of a mother logical volume.}
\label{fig:overlap}
\end{center}
\end{figure}

%\begin{algorithm}[h]
%\SetAlgoLined
%\KwResult{Write here the result }
% initialization\;
% \While{While condition}{
%  instructions\;
%  \eIf{condition}{
%   instructions1\;
%   instructions2\;
%   }{
%   instructions3\;
%  }
% }
% \caption{How to write algorithms}
%\end{algorithm}

@@

For a logical volume with $n_{\rm daughters}$ physical volumes, assuming meshes for solids with average number of vertices $n$, clearly this algorithm 
has complexity $\sim \mathcal{O}(n^2 n_{\rm daughters})$. This might seem like a prohibitive computational cost, but worth it considering the potential 
waste if small overlaps are present in the final RTMC simulation. This algorithm clearly favours geometry descriptions which have a high degree of logical 
volume reuse. Due to the discrete nature of triangular meshes it is not possible to have perfect detection of overlaps, especially when curved surfaces are 
considered.

@@

\section{GDML to FLUKA conversion}
It is relatively straight forward to convert Geant4 geometry to FLUKA. Each of the Geant4 
solids can be mapped to a FLUKA \verb|REGION|. Some solids in Geant4 directly map to 
a FLUKA body, others need a construction of a simple FLUKA CSG tree. Table~\ref{tab:geant2fluka}
lists the mapping between Geant4 solids and the bodies used to compose a FLUKA region.
A region is a region of space which is formed of bodies and the operators + (intersection), - 
(complement) and | (union), also parentheses to indicate the order of operations. FLUKA 
has XXX bodies and in general infinite extent bodies have benefits over finite ones. Key 
for conversion are XY-, XZ-, YZ-Planes (XYP, XZP, YZP), arbitrary plane (PLA), 
X, Y,  Z axis aligned cylinders (XCC, YCC, ZCC), X, Y, Z axis aligned elliptical cylinders (XEC, YEC, ZEC), 
sphere (SPH), truncated right-angle cone (TRC) and general quadric surface (QUA).

@@

Figure~\ref{fig:gdml-flair} shows an example conversion from GDML to FLUKA. The example is a 
vacuum chamber with three ConFlat flange (CF) beam pipes connected to CF flanges. The top and bottom plates 
have been removed to display the geometry more clearly. The model is formed of \verb|G4Box| and
\verb|G4Tubs| and Boolean operations of subtraction, intersection and union.
\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=1.0\textwidth]{./model-scene/CuboidalChamber.pdf}
\caption{Example conversion of a simple GDML geometry to FLUKA, left panel:  
the original model in \PYGEOMETRY{}, right panel: the converted FLUKA geometry 
viewed in FLAIR. The example is a sector bend dipole electromagnet.}
\label{fig:gdml-flair}
\end{center}
\end{figure*}

@@

\begin{enumerate}
\item Degenerate body look up 
\item Bodies with/without transformation 
\item Non-convex solids, convex decomposition (TwistedBox, TwistedTrap, TwistedTrd, TwistedTube, Tessellated)
\item Scaled solid, Placement scale 
\item Replica, Division, Parameterised
\item Recursive application of \ref{eqn:setDiff}, \ref{eqn:setUnion}, \ref{eqn:setIntersection}, \ref{eqn:logicalSubtraction} speed up using meshes
\end{enumerate}

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=1.0\textwidth]{./model-scene/SectorBend.pdf}
\caption{Example conversion of a simple FLUKA geometry to GDML, left panel:  
the original model in FreeCAD, right panel: the GDML geometry 
viewed in \PYGEOMETRY{}. The example is a sector bend dipole electromagnet.}
\label{fig:cad-gdml}
\end{center}
\end{figure*}

@@

\begin{enumerate}
\item Graphical front end viewer
\item C++ output
\item CAD BREP to FLUKA converter (a la McCAD) 
\item Integration of other tools (DAGMC) 
\item Far from perfect conversions
\item Export mesh for other applications (GPU optics)
\item Convergence FLUKA/Geant4 - Moira 
\item Replace G4 polyhedron CSG
\item Not for large scale conversion
\end{enumerate}

>>>>

There are existing codes that have functionality similar to \PYGEOMETRY{}. ROOT \cite{Brun:1997pa}, the high-energy physics data analysis program, can load, display and manipulate GDML like geometry. Tools exist to convert CAD files to GDML, open source examples are GUIMesh \cite{GUIMesh_Pinto}, commercial products include ESABASE2 \cite{ESABASE2} and FASTRAD \cite{FASTRAD}. There are also tools from the fusion and 
neutronics community which  can convert CAD geometry into formats usable by RTMC codes, examples include DAGMC \cite{DAGMC} and McCad \cite{McCad}.  
In principle CAD software can export shape data to STL (or other similar mesh formats), which can be used by Geant4 \cite{poole2012acad}. In practice using a lot of CAD models is difficult if a CAD model comprises of a large number of parts, exporting, assigning material to and placing the STL components into the RTMC can be a 
cumbersome task. The existent set of software does not provide a complete set of tools to efficiently create complex geometries.

@@

%%%%%%%%%%%%%%%%% Geant4 geometery
Geant4 geometry description is a mixture of BREP, CSG and tessellated concepts. Geant4 includes 27 basic solids, although it does not store a sense 
of topology present in traditional CAD BREP systems. One of the fundamental solids is a {\em TessellatedSolid} which can used to represent STL or PLY files. 
Geant4 also provides the ability to perform Boolean operations on these primitive solids. The richest and most flexible geometry description is currently used by
Geant4. Not only do solid objects need to be defined but also placed in a world coordinate system. Geant4 has two concepts which facilitate this {\it Logical Volumes} and {\it Physical Volumes}. A {\it Logical Volume} is a region of space  that is defined by an outer solid but also other attributes like material, magnetic field 
and zero or more physical {\it daughter} volumes. A Physical Volume is a placement (or instance) of a Logical Volume.

@@

\subsection{Tessellation of solids (meshing)}
Creating a uniform three dimensional mesh description of all solids (include Booleans) is exceptionally useful, for visualisation and other algorithms, e.g overlap 
detection. For each Geant4 solid instance a triangular tessellated Vertex-Face mesh is generated and cached. This mesh is then used to determine the extent 
of placed instances of geometry (physical volumes) and meshes for CSG derived solids. CSG mesh calculations are performed using a BSP tree technique in 
pure PYTHON \cite{pycsg} or the Computational Geometry Algorithms Library (CGAL) surface meshes \cite{cgal:bsmf-sm-20b}. In general the CGAL implementation is one or two orders of magnitude faster than the pure-PYTHON CSG implementation and must be used for large geometries. Triangular meshes based on CSG 
operations involving curved surfaces often contain large numbers of thin triangles. Before meshes are visualised or written to disk various polygon mesh 
algorithms  from CGAL \cite{cgal:lty-pmp-20b} can be employed to give the meshes more desirable features.

@@

\subsection{Visualisation} \label{sec:visualisation}
When implementing geometry a rapid and robust visualisation system is key to produce error free and efficient simulation input.
A  \PYGEOMETRY{}  hierarchy can be viewed using the popular Visualisation Toolkit (VTK). No separate scene graph is required as the Geant4 volume hierarchy is sufficient 
to place the meshes associated with each physical volume. A daughter volume is placed within a logical volume with a rotation $\mathbf{R}_d$, scale $\mathbf{S}_d$ and translation $\mathbf{T}_d$.

@@

\subsection{Overlap detection}
All RTMC codes can not handle spatial overlap between two geometric objects and likely are to have ill-defined behaviour when tracking particles  
through such a situation.  A key feature of \PYGEOMETRY{} is the detection of potential overlaps in a way which is most useful to the user, it does this by performing an intersection operation between solid instances and determining if the resulting mesh is empty. Figure~\ref{fig:overlap} shows the potential overlaps, a) protrusion of a daughter from the mother, b)  finite volume intersection between two daughters  and c) an overlap where two daughters share a face. If the resulting intersection 
is non-null then the overlaps can be displayed side-by-side in the visualisation. 
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=8cm]{./diagrams/overlap.pdf}
\caption{Schematic of the three different types of overlap between daughters of a mother logical volume.}
\label{fig:overlap}
\end{center}
\end{figure}

\begin{algorithm}[h]
\SetAlgoLined
\KwResult{{Write here the result} }
 initialization\;
 \While{While condition}{
  instructions\;
  \eIf{condition}{
   instructions1\;
   instructions2\;
   }{
   instructions3\;
  }
 }
 \caption{How to write algorithms}
\end{algorithm}

@@

For a logical volume with $n_{\rm daughters}$ physical volumes, assuming meshes for solids have an average number of faces $n$, clearly this algorithm 
has complexity $\sim \mathcal{O}(n^2 n_{\rm daughters}^2)$. This might seem like a prohibitive computational cost, but worth it considering the potential 
waste if small overlaps are present in the final RTMC simulation and large amounts of cluster CPU time is wasted. This algorithm clearly favours geometry descriptions which have a high degree of logical volume reuse. Due to the discrete nature of triangular meshes it is not possible to have perfect detection 
of overlaps, especially when curved surfaces are considered.

@@

\section{GDML to FLUKA conversion}
It is relatively straight forward to convert Geant4 geometry to FLUKA. Each of the Geant4 
solids can be mapped to a FLUKA \verb|REGION|. Some solids in Geant4 directly map to 
a FLUKA body, others need a construction of a simple FLUKA CSG tree. Table~\ref{tab:geant2fluka}
lists the mapping between Geant4 solids and the bodies used to compose a FLUKA region.
A region is a region of space which is formed of bodies and the operators + (intersection), - 
(complement) and | (union), also parentheses to indicate the order of operations. FLUKA 
has 20 bodies, listed in Table~\ref{tab:Fluka2Geant4}  and in general infinite extent bodies have benefits over finite ones. Key 
for conversion are XY-, XZ-, YZ-Planes (XYP, XZP, YZP), arbitrary plane (PLA), 
Z axis aligned cylinder (ZCC), Z axis aligned elliptical cylinder (ZEC), 
sphere (SPH), truncated right-angle cone (TRC) and general quadric surface (QUA).

@@

Figure~\ref{fig:gdml-flair} shows an example conversion from GDML to FLUKA. The example is a 
vacuum chamber with three ConFlat flange (CF) beam pipes connected to CF flanges. The top and bottom plates 
have been removed to display the geometry more clearly. The model is formed of \verb|G4Box| and
\verb|G4Tubs| and Boolean operations of subtraction, intersection and union.
\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{./model-scene/CuboidalChamber.pdf}
\caption{Example conversion of a simple GDML geometry to FLUKA, left panel:  
the original model in \PYGEOMETRY{}, right panel: the converted FLUKA geometry 
viewed in FLAIR. The example is a sector bend dipole electromagnet.}
\label{fig:gdml-flair}
\end{center}
\end{figure*}

@@

\begin{enumerate}
\item Degenerate body look up 
\item Bodies with/without transformation 
\item Non-convex solids, convex decomposition (TwistedBox, TwistedTrap, TwistedTrd, TwistedTube, Tessellated)
\item Scaled solid, Placement scale 
\item Replica, Division, Parameterised
\item Recursive application of \ref{eqn:setDiff}, \ref{eqn:setUnion}, \ref{eqn:setIntersection}, \ref{eqn:logicalSubtraction} speed up using meshes
\item Issues with tessellated solids 
\end{enumerate}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{./model-scene/SectorBend.pdf}
\caption{Example conversion of a simple FLUKA geometry to GDML, left panel:  
the original model in FreeCAD, right panel: the GDML geometry 
viewed in \PYGEOMETRY{}. The example is a sector bend dipole electromagnet.}
\label{fig:cad-gdml}
\end{center}
\end{figure*}

@@

\begin{enumerate}
\item Graphical front end viewer
\item C++ output
\item CAD BREP to FLUKA converter (a la McCAD) 
\item Integration of other tools (DAGMC) 
\item Far from perfect conversions
\item Export mesh for other applications (GPU optics)
\item Convergence FLUKA/Geant4 - Moira 
\item Replace G4 polyhedron CSG
\item Not for large scale conversion
\item Export for other RTMC codes (MCNP, etc) 
\end{enumerate}
